\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-26 coqimpl.tex
Rewriting global reasoning.
Add example for global reasoning and MetaData. Reply comments. Add "FieldClaim" info

% explain ftheorem wrapping this complicated recursor
We also provide a command "FTheorem" when the users want to use tactic
programming instead. This tool can avoid most boilerplate code (e.g.,
specify the handler family) when writing a recursor like above, invoke
proof interaction mode, allow tactic usage, and is also open to
extension like "FRecursor".\YZ{example?}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases the users just want global reasoning instead of an extensible proof. For example, in STLC, we should expect`` \mintinline{Coq}{~ value (tm_app x y)}'' holds in all possible extension of STLC (i.e. application is never a terminal value). We know the correct way of proving this proposition \mintinline{Coq}{Field app_not_value : value (tm_app x y) -> False.} is to use "FRecursor" to inductively reason "value (tm_app x y)". However, this means that everytime we extend "value" with new terminal value, we will need to extend corresponding inductive cases for "app_not_value". We can imagine the newly extended proof to be boring case analysis that the newly extended terminal value is not "tm_app". What's worse, similar scenario can happen on other statement like \mintinline{Coq}{~ value (tm_if cond x y)}. Everything here is extended correctly but we end up having boilerplate code, \textbf{semantically}.

Our proposal is to allow global reasoning, which will look like a constraint inside the family. We provide "Closing Fact" as a command to carry out this global reasoning, attached with a proof script and proof term. This "Closing Fact" will look like an assertion during the construction of the surrounding family, and this assertion will only be verified when the whole family is closed and compiled into a module, with the help of the attached proof script and proof terms.

For example,

@@

@@

% This style of ``Global Reasoning'' says that we can prove the
% corresponding theorem only when we know the ``full picture'', in particular,
% all the constructors of an inductive type. 
% We provide "Closing Fact" as
% a command to carry out this global reasoning
\YZ{Are recursors and partial recursors automatically generated and proved by the plugin, or must they be asserted and proved by the programmer?}\EDJreply{(1) Closing Fact is not only used to generate recursor and partial recursor. I think the intro of this section is just giving people wrong impression. So I rewrite the whole sectio now. (2)  Recursors and Partial recursors are possible to be automatically generated but my plugin didn't implement it. Current stage is that user only need to assert them. Nobody has to prove it (plugin will prove it). Here when I say user has to provide proof script, it means for arbitrary proposition if the user want to carry out global reasoning.}
% ---we can attach to "Closing
% Fact" either (proof) terms or Coq's proof script, then only when
% compilation to module happens, is the term/proof script
% type-checked to see if a "Closing Fact" statement can be proved. Note
% that, "Closing Fact" cannot immediately check the correctness of the
% term/proof script because this verification only happens when we
% finished defining a family, and thus a bit harder to handle.

% This "Closing Fact" also can ease us from the necessity to let plugin
% generate proof for recursors and partial recursors and their
% computational rules (as in \ref{langdesign-req4}), and thus making
% plugin development easier.
\YZ{This reads like a bad excuse: the programmer does not care about if the plugin development is easy. Question is if it makes the programmer's life easier.}\EDJreply{You are right. It does sound like a bad excuse. I rewrite the whole section now. Closing Fact is initially used when the user wants global reasoning instead of extensible proof by induction, for example, the inversion lemma. \\ It just happens that they can be used by me so that I can postpone generation of these rule plugin (which is hard for me to do). This paragraph is just me being honest that my plugin is not mature enough to generate everything.}
% For example, the computation rule for a
% concrete recursor like "B.neg" and partial recursors in
% \cref{fig:plugin-example2} can be directly asserted by using "Closing
% Fact" and their proofs verified later during compilation.
\YZ{What is the alternative if not using "Closing Fact"?}\EDJreply{To be a legit plugin, my plugin should generate them...}
% This ``Global Reasoning'' will be again useful in later examples.\YZ{Still not entirely clear to me what is gained and what is lost by using "Closing Fact".}

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family,
the family data is invisible to Coq internal and thus causes some
inconvenience.  For example, currently it is not possible to define a
vanilla inductive type dependent on our ``extensible'' inductive type
due to the invisibility, unless we compile the whole family. Similarly,
customized tactic expressions as required by \ref{langdesign-req5} cannot
use lemmas or theorems in the fields of the current defining family as
well.

To remedy this, we support a "MetaData" command. This command
bundles any original Coq primitive (in the current family
context) into the compiled module type, and makes them visible to the
following defining field. Doing so, we can define customized tactic as
in \ref{langdesign-req5} that uses lemmas proved in the family.\YZ{Consider adding examples of Closing Fact and MetaData, even if only to be included in an appendix.}

>>>>

% explain ftheorem wrapping this complicated recursor
We also provide a command "FTheorem" when the users want to use tactic
programming instead. This tool can avoid most boilerplate code (e.g.,
specify the handler family) when writing a recursor like above, invoke
proof interaction mode, allow tactic usage, and is also open to
extension like "FRecursor".\YZ{example?}\EDJreply{FTheorem is really just a short hand for "FRecursor" + handler family. Nothing more conceptual here. If the reader is interested in the concrete syntax, I think I need to make the two examples avaliable in the appendix.}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases the users just want global reasoning instead of an extensible proof. For example, in STLC, we should expect`` \mintinline{Coq}{~ value (tm_app x y)}'' holds in all possible extension of STLC (i.e. application is never a terminal value). We know the correct way of proving this proposition \mintinline{Coq}{value_not_app : value (tm_app x y) -> False.} is to use "FRecursor" to inductively reason "value (tm_app x y)". However, this means that everytime we extend "value" with new terminal value, we will need to extend corresponding inductive cases for "app_not_value". We can imagine the newly extended proof to be boring case analysis that the newly extended terminal value is not "tm_app". What's worse, similar scenario can happen on other statement like \mintinline{Coq}{~ value (tm_if cond x y)}. Everything here is extended correctly but we end up having boilerplate code, \textbf{semantically}.
\begin{figure}[!htb]
\begin{minted}{Coq}
Family STLC.
  FInductive tm : Set := ... 
  FInductive value : tm -> Prop := ... 
  Closing Fact value_not_app : forall x y, ~ value (tm_app x y) 
      by { intros x y H; inversion H; eauto }.
  Fail Ltac inv := 
    match goal with 
    | [h : value (tm_app _ _) |- ] => destruct (value_not_app _ _ h)
    (* Fail: value, tm_app, value_not_app unfound *)
    end. 
  MetaData tactic1.
    Ltac inv := (* ... the same definition as above ... *)
  EndMetaData.
EndFamily.
\end{minted}  
\caption{Example for Global Reasoning and MetaData}\label{fig:plugin-example-global-reasoning-meta-data}
\end{figure}

Our proposal is to allow global reasoning, which will look like a constraint inside the family. We provide "Closing Fact" as a command to carry out this global reasoning, attached with a proof script and proof term. This "Closing Fact" will look like an assertion during the construction of the surrounding family. This assertion will only be verified when the whole family is closed and compiled into a module, with the help of the attached proof script and proof terms.

Here we show the example to prove "value_not_app" in \cref{fig:plugin-example-global-reasoning-meta-data}, the script inside curly bracket after "by" is the attached proof script that will be executed when the whole family is being compiled. Since we are dealing with a concrete inductive type during compilation, we can use inversion tactic on the vanilla inductive type.  This "Closing Fact", as an assertion, will have no problem being inherited but the proof script will be executed everytime during each compilation of the children families.

@@

Ultimately, "Closing Fact" is asserting a constraint upon all the possible extension 
of the surrounding family and claiming a proof script can solve the constraints. 
We expect it to be used in a light-weighted form for maintainability.

@@

% This style of ``Global Reasoning'' says that we can prove the
% corresponding theorem only when we know the ``full picture'', in particular,
% all the constructors of an inductive type. 
% We provide "Closing Fact" as
% a command to carry out this global reasoning
\YZ{Are recursors and partial recursors automatically generated and proved by the plugin, or must they be asserted and proved by the programmer?}\EDJreply{(1) Closing Fact is not only used to generate recursor and partial recursor. I think the intro of this section is just giving people wrong impression. So I rewrite the whole section now. (2)  Recursors and Partial recursors are possible to be automatically generated but my plugin didn't implement it. Current stage is that user only need to assert them. Nobody has to prove it (plugin will prove it). Here when I say the user has to provide proof script, it means for arbitrary proposition if the user want to carry out global reasoning.}
% ---we can attach to "Closing
% Fact" either (proof) terms or Coq's proof script, then only when
% compilation to module happens, is the term/proof script
% type-checked to see if a "Closing Fact" statement can be proved. Note
% that, "Closing Fact" cannot immediately check the correctness of the
% term/proof script because this verification only happens when we
% finished defining a family, and thus a bit harder to handle.

% This "Closing Fact" also can ease us from the necessity to let plugin
% generate proof for recursors and partial recursors and their
% computational rules (as in \ref{langdesign-req4}), and thus making
% plugin development easier.
\YZ{This reads like a bad excuse: the programmer does not care about if the plugin development is easy. Question is if it makes the programmer's life easier.}\EDJreply{You are right. It does sound like a bad excuse. I rewrite the whole section now. Closing Fact is initially used when the user wants global reasoning instead of extensible proof by induction, for example, the inversion lemma. \\ It just happens that they can be used by me so that I can postpone generation of these rule plugin (which is hard for me to do). This paragraph is just me being honest that my plugin is not mature enough to generate everything.}
% For example, the computation rule for a
% concrete recursor like "B.neg" and partial recursors in
% \cref{fig:plugin-example2} can be directly asserted by using "Closing
% Fact" and their proofs verified later during compilation.
\YZ{What is the alternative if not using "Closing Fact"?}\EDJreply{(1) For partial recursor and recursor, that should be generated by plugin (2) For general global reasoning, I don't know. There must be some other ways in the literature. }
% This ``Global Reasoning'' will be again useful in later examples.\YZ{Still not entirely clear to me what is gained and what is lost by using "Closing Fact".}

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family,
the family data is invisible to Coq internal and thus causes some
inconvenience. For example, in \cref{fig:plugin-example-global-reasoning-meta-data} the first attempt of defining tactic inv fails. Because the earlier defining "tm", "value" and "value_not_app" are currently just data structure in the plugin and not visible for Coq internal (only visible after the family is compiled into module).

Thus to refer to them, we have to scope the defining tactic with "MetaData" and our plugin will setup the appropriate environment. After a "MetaData" block is closed, the plugin will aggregate the content into the current defining family so that they can also be used when defining the following fields of the family. This can fulfill \ref{langdesign-req5}.

To fulfill \ref{langdesign-req5}, we need somehow to aggregate tactic inside compiled module type, due to the implementation of our family---when defining every field of

% For example,

%   For example, currently it is not possible to define a
% vanilla inductive type dependent on our ``extensible'' inductive type
% due to the invisibility, unless we compile the whole family. Similarly,
% customized tactic expressions as required by \ref{langdesign-req5} cannot
% use lemmas or theorems in the fields of the current defining family as
% well.

% To remedy this, we support a "MetaData" command. This command
% bundles any original Coq primitive (in the current family
% context) into the compiled module type, and makes them visible to the
% following defining field. Doing so, we can define customized tactic as
% in \ref{langdesign-req5} that uses lemmas proved in the family.\YZ{Consider adding examples of Closing Fact and MetaData, even if only to be included in an appendix.}

### 2022-06-26 lang-design.tex
Add example for global reasoning and MetaData. Reply comments. Add "FieldClaim" info

\subsection{Language Design}
Inspired by our above requirements and Family Polymorphism, we propose
the following extension to Coq's Vernacular Commands.
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle \quad | \quad ...
\end{align*}

These are some core extension we proposed to support \cref{fig:STLC-example}. We use "Family" and "EndFamily" to scope, and "Field" is used to declare new members in a family.
We can define nested families.
Family will also be used to encode recursive handlers. Each case
handling of one inductive type will be aggregated into one family, then
"FRecursor" will be used to construct a recursor using these handlers.
For example, \mintinline{Coq}{FRecursor recur about T motive f} "using"
"handlers" will generate a new field "recur" that is doing induction
(using elimination principle of "T") using motive "f", and the handlers
are located in the family "handlers".

>>>>

\subsection{Language Design}
Inspired by our above requirements and Family Polymorphism, we propose
the following extension to Coq's Vernacular Commands.
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "FieldClaim"~\langle ident\rangle~":"~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle \quad | \quad ...
\end{align*}

These are some core extension we proposed to support \cref{fig:STLC-example}. We use "Family" and "EndFamily" to scope, and "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.}
We can define nested families.
Family will also be used to encode recursive handlers. Each case
handling of one inductive type will be aggregated into one family, then
"FRecursor" will be used to construct a recursor using these handlers.
For example, \mintinline{Coq}{FRecursor recur about T motive f} "using"
"handlers" will generate a new field "recur" that is doing induction
(using elimination principle of "T") using motive "f", and the handlers
are located in the family "handlers".
